quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-11-08 paper-quollr.Rmd
Changed the notations to view in HTML as well

---
title: "quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-linear dimension reduction (NLDR) methods provide a low-dimensional representation of high-dimensional data ($p\text{-}D$) by applying a non-linear transformation. However, the complexity of the transformations and data structures can create wildly different representations depending on the method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures. The R package \CRANpkg{quollr} has been developed as a new visual tool to determine which method and which hyper-parameter choices provide the most accurate representation of high-dimensional data. The `scurve` data from the package is used to illustrate the algorithm. Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles are used to demonstrate the usability of the package.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
  \newcommand{\pD}{$p\text{-}D$}
  \newcommand{\twoD}{$2\text{-}D$}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-quollr.bib
editor_options: 
  chunk_output_type: inline
---

@@

This paper presents the R package, `quollr`, which is useful for understanding how NLDR warps high-dimensional space and fits the data. Starting with an NLDR layout, our approach is to create a \twoD wireframe model representation, that can be lifted and displayed in the high-dimensional (\pD{}) space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:overviewhtml)"} else { "Figure \\@ref(fig:overview)"})`).

@@

```{r overviewhtml, eval=knitr::is_html_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different \\twoD projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H', layout = "l-page"}

@@

```{r overview, eval=knitr::is_latex_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different \\twoD projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H'}

@@

The package is organized into seven core modules corresponding to stages of the analysis workflow: preprocessing, \twoD model construction, lifting into \pD, prediction, error computation, visualization, and interactivity. Each module performs a distinct task and communicates through data objects.

@@

2. Construct \twoD model: A series of functions `hex_binning()`, `merge_hexbin_centroids()`, `tri_bin_centroids()`, `gen_edges()`, and `update_trimesh_index()` generate the hexagonal grid, compute bin centroids, and connect the triangular mesh that defines local neighborhoods in the \twoD space.

3. Lift the model into \pD: The function `avg_highd_data()` computes the average of the high-dimensional variables for each bin, linking the \twoD representation back to the original data space.

@@

7. Interactivity: The functions `comb_all_data_model()` and `show_link_plots()` generate interactive linked visualizations that connect the \twoD NLDR layout, the corresponding tour view, and the fitted model. Similarly, `comb_all_data_model_error()` and `show_error_link_plots()` integrate the error distribution with the \twoD embedding and dynamic high-dimensional view, enabling interactivity across multiple visual components.

@@

- `highd_vis_model`: the main model object returned by `fit_highd_model()`. It stores all components of the fitted visualization model, including the scaled NLDR data (`nldr_scaled_obj`), the hexagonal bin structure (`hb_obj`), the averaged \pD summaries for each bin (`model_highd`), the corresponding \twoD bin centroids (`model_2d`), and the triangulated mesh connecting neighboring bins (`trimesh_data`).

@@

Several core computations within `quollr` are optimized using compiled C++ code via the `Rcpp` and `RcppArmadillo` packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (`compute_highd_dist()`), error metrics (`compute_errors()`), \twoD distance calculations (`calc_2d_dist_cpp()`), and generation of hexagon coordinates (`gen_hex_coord_cpp()`) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`hd_thresh`). The function returns an object of class `highd_vis_model` containing the scaled NLDR object (`nldr_scaled_obj`) with three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both \twoD (`model_2d`), and \pD (`model_highd`), and triangular mesh (`trimesh_data`).

@@

Constructing the \twoD model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

Finally, a tibble is returned containing a unique hexagon ID (`h`) along with the corresponding x and y centroid coordinates (`c_x`, `c_y`), which define the layout of the hexagonal grid over the \twoD space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The `assign_data()` function performs this assignment by calculating the \twoD Euclidean distance between each point in the \twoD embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the \twoD layout. It then selects the corresponding x and y coordinates of each hexagon’s centroid.

@@

The result is a `tibble` of the scaled \twoD embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by `h`, which represents the hexagon ID associated with each point in the \twoD layout. Within each group, it collects the `ID`s into a list, resulting in a summary where each row corresponds to a single hexagon. The resulting column, `pts_list`, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a \twoD layout, we employ Delaunay triangulation [@lee1980;@albrecht2024] on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or "holes" in the \twoD structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted \twoD model into \pD. This is done by modelling a point in \pD as the  \pD mean of data points in the \twoD centroid. This is performed using the `avg_highd_data()` function, which takes \pD data (`highd_data`) and embedding data with their corresponding hexagonal bin IDs as inputs (`scaled_nldr_hexid`).

@@

The `predict_emb()` function is used to predict a point in a \twoD embedding for a new \pD data point using the fitted model. This function is useful to predict \twoD embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest \pD model point is identified for the new \pD data point by computing \pD Euclidean distance. Then, the corresponding \twoD bin centroid mapping for the identified \pD model point is determined. Finally, the coordinates of the identified \twoD bin centroid is used as the predicted NLDR embedding for the new \pD data point.

@@

## Compute residuals and Root Within Bin Sum of Square (RWBSS)

Root Within Bin Sum of Square (RWBSS) are used as goodness of fit metrics for the model. These metrics can be computed using the  `glance()` function, which provides a tidy output for evaluation.

The function requires both the fitted model object returned by `fit_highd_model()` and \pD data to begin. The \pD model output (`model_highd`) is first renamed to avoid naming conflicts during subsequent data joins. It then uses the `predict_emb()` function to assign each point in the \pD dataset to a corresponding hexagon bin in the \twoD model, producing a prediction data frame that contains both the predicted bin assignment (`pred_h`) and the original observation `ID`.

The function joins this prediction output with both the \pD model and the \pD data (to retrieve true coordinates). It then calculates squared differences between the original and predicted \pD coordinates for each dimension, storing these as `error_square_x1`, `error_square_x2`, ..., up to the dimensionality of the data.

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the RWBSS which is the average of the total squared error per point.

These metrics are returned in a tibble as `Error` (absolute error) and `RWBSS` (root mean squared error). The computation of total absolute error and RWBSS is performed in C++ for efficiency using the internal `compute_errors()` function.

@@

Furthermore, `augment()` requires both the fitted model object returned by `fit_highd_model()` and \pD data to begin. It extends the fitted model by adding prediction results and error diagnostics to the original \pD data.

The function starts with the same process as is used in the `glance()` function to produce a predicted point in \pD for each point in the \pD dataset.

@@

The package offers several \twoD visualizations, including:

@@

The generated \pD model, overlaid with the data, can also be visualized using `show_langevitour`. Additionally, it features a function for visualizing the \twoD projection of the fitted model overlaid on the data, called `plot_proj`.

@@

The `geom_trimesh()` function introduces a custom `ggplot2` layer designed for visualizing \twoD wireframe on a provided set of bin centroids.

@@

### \pD model visualization

To visualize how well the \pD model captures the underlying structure of the high-dimensional data, we provide a tour of the model in \pD using the `show_langevitour()` function. This function renders a dynamic projection of both the high-dimensional data and the model using the `langevitour` R package [@paul2023].

@@

As an alternative to `langevitour`, users can explore the fitted \pD model using the `detourr` [@casper2025]. The combined data object from `comb_data_model()` can be passed directly to the `detour()` function, where `tour_aes()` defines the projection variables and color mapping. The visualization is rendered using `show_scatter()`, which can display both data points and the model’s structural edges via the `edges` argument.

@@

In the resulting interactive visualization, blue points represent the high-dimensional data, orange points represent the model centroids from each bin, and the lines between model points reflect the \twoD wireframe structure mapped to high-dimensional space.

@@

To support interactive evaluation of how well the \pD model captures the structure of the high-dimensional data, we introduce `show_link_plots()`. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the \twoD bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a \twoD NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), \twoD model data (`model_2d`), and model error (`error_data`).

@@

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a \twoD NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

As an alternative to using `langevitour`, link plots can also be generated with the `detourr` package. In this setup, users can manually construct the linked visualization using the `crosstalk` [@joe2023] and `htmltools` [@joe2024] packages. The interactive layout is created by arranging the \twoD NLDR plot (`nldr_plt`), the optional error distribution plot (`error_plt`), and the tour view produced with `detourr`, side by side within a flexible grid. The NLDR and error plots are rendered with `ggplotly()` [@chapman2020] to enable interactive linking. The `bscols()` function from `crosstalk` manages synchronization across these panels, allowing for linked brushing and coordinated selection between these interactive plots.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rwbss) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \twoD layout.

@@

error_limb <- error_limb |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  filter(a1 >= 0.03) |>
  group_by(method, a1) |>
  filter(RMSE == min(RMSE)) |>
  ungroup()

@@

error_plot_limb <- plot_rmse(error_limb) +
  scale_x_continuous(breaks = sort(unique(error_limb$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#984ea3','#4daf4a','#ff7f00'))

@@

```{r limb-rwbss, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RWBSS for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\\label{fig:limb-rwbss}", fig.pos='H', echo=TRUE}

@@

plot_rmse_layouts(plots = list(error_plot_limb, nldr1, 
                               nldr2, nldr3, nldr4, 
                               nldr5, nldr6), design = design)
```

@@

```{r model-limb, fig.cap="Compare the published $2\\text{-}D$ layout (Figure \\ref{fig:limb-rwbss}b) and the $2\\text{-}D$ layout selected (Figure \\ref{fig:limb-rwbss}f) by RWBSS plot (Figure \\ref{fig:limb-rwbss}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", fig.pos='H', fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

@@

The `quollr` package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in \twoD and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a \twoD layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate \twoD representations.

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, `quollr` provides a quantitative route for evaluation. It enables the computation of RWBSS and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the `langevitour` package, allowing users to inspect where the model fits well and where it does not.

>>>>

---
title: "quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-linear dimension reduction (NLDR) methods provide a low-dimensional representation of high-dimensional data ($p\text{-}D$) by applying a non-linear transformation. However, the complexity of the transformations and data structures can create wildly different representations depending on the method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures. The R package \CRANpkg{quollr} has been developed as a new visual tool to determine which method and which hyper-parameter choices provide the most accurate representation of high-dimensional data. The `scurve` data from the package is used to illustrate the algorithm. Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles are used to demonstrate the usability of the package.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-quollr.bib
editor_options: 
  chunk_output_type: inline
---

@@

This paper presents the R package, `quollr`, which is useful for understanding how NLDR warps high-dimensional space and fits the data. Starting with an NLDR layout, our approach is to create a $2\text{-}D$ wireframe model representation, that can be lifted and displayed in the high-dimensional ($p\text{-}D$) space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:overviewhtml)"} else { "Figure \\@ref(fig:overview)"})`).

@@

```{r overviewhtml, eval=knitr::is_html_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\text{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H', layout = "l-page"}

@@

```{r overview, eval=knitr::is_latex_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\text{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H'}

@@

The package is organized into seven core modules corresponding to stages of the analysis workflow: preprocessing, $2\text{-}D$ model construction, lifting into $p\text{-}D$, prediction, error computation, visualization, and interactivity. Each module performs a distinct task and communicates through data objects.

@@

2. Construct $2\text{-}D$ model: A series of functions `hex_binning()`, `merge_hexbin_centroids()`, `tri_bin_centroids()`, `gen_edges()`, and `update_trimesh_index()` generate the hexagonal grid, compute bin centroids, and connect the triangular mesh that defines local neighborhoods in the $2\text{-}D$ space.

3. Lift the model into $p\text{-}D$: The function `avg_highd_data()` computes the average of the high-dimensional variables for each bin, linking the $2\text{-}D$ representation back to the original data space.

@@

7. Interactivity: The functions `comb_all_data_model()` and `show_link_plots()` generate interactive linked visualizations that connect the $2\text{-}D$ NLDR layout, the corresponding tour view, and the fitted model. Similarly, `comb_all_data_model_error()` and `show_error_link_plots()` integrate the error distribution with the $2\text{-}D$ embedding and dynamic high-dimensional view, enabling interactivity across multiple visual components.

@@

- `highd_vis_model`: the main model object returned by `fit_highd_model()`. It stores all components of the fitted visualization model, including the scaled NLDR data (`nldr_scaled_obj`), the hexagonal bin structure (`hb_obj`), the averaged $p\text{-}D$ summaries for each bin (`model_highd`), the corresponding $2\text{-}D$ bin centroids (`model_2d`), and the triangulated mesh connecting neighboring bins (`trimesh_data`).

@@

Several core computations within `quollr` are optimized using compiled C++ code via the `Rcpp` and `RcppArmadillo` packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (`compute_highd_dist()`), error metrics (`compute_errors()`), $2\text{-}D$ distance calculations (`calc_2d_dist_cpp()`), and generation of hexagon coordinates (`gen_hex_coord_cpp()`) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`hd_thresh`). The function returns an object of class `highd_vis_model` containing the scaled NLDR object (`nldr_scaled_obj`) with three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both $2\text{-}D$ (`model_2d`), and $p\text{-}D$ (`model_highd`), and triangular mesh (`trimesh_data`).

@@

Constructing the $2\text{-}D$ model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

Finally, a tibble is returned containing a unique hexagon ID (`h`) along with the corresponding x and y centroid coordinates (`c_x`, `c_y`), which define the layout of the hexagonal grid over the $2\text{-}D$ space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The `assign_data()` function performs this assignment by calculating the $2\text{-}D$ Euclidean distance between each point in the $2\text{-}D$ embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the $2\text{-}D$ layout. It then selects the corresponding x and y coordinates of each hexagon’s centroid.

@@

The result is a `tibble` of the scaled $2\text{-}D$ embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by `h`, which represents the hexagon ID associated with each point in the $2\text{-}D$ layout. Within each group, it collects the `ID`s into a list, resulting in a summary where each row corresponds to a single hexagon. The resulting column, `pts_list`, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation [@lee1980;@albrecht2024] on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or "holes" in the $2\text{-}D$ structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted $2\text{-}D$ model into $p\text{-}D$. This is done by modelling a point in $p\text{-}D$ as the  $p\text{-}D$ mean of data points in the $2\text{-}D$ centroid. This is performed using the `avg_highd_data()` function, which takes $p\text{-}D$ data (`highd_data`) and embedding data with their corresponding hexagonal bin IDs as inputs (`scaled_nldr_hexid`).

@@

The `predict_emb()` function is used to predict a point in a $2\text{-}D$ embedding for a new $p\text{-}D$ data point using the fitted model. This function is useful to predict $2\text{-}D$ embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest $p\text{-}D$ model point is identified for the new $p\text{-}D$ data point by computing $p\text{-}D$ Euclidean distance. Then, the corresponding $2\text{-}D$ bin centroid mapping for the identified $p\text{-}D$ model point is determined. Finally, the coordinates of the identified $2\text{-}D$ bin centroid is used as the predicted NLDR embedding for the new $p\text{-}D$ data point.

@@

## Compute residuals and hexbin error (HBE)

Hexbin error (HBE) are used as goodness of fit metrics for the model. These metrics can be computed using the  `glance()` function, which provides a tidy output for evaluation.

The function requires both the fitted model object returned by `fit_highd_model()` and $p\text{-}D$ data to begin. The $p\text{-}D$ model output (`model_highd`) is first renamed to avoid naming conflicts during subsequent data joins. It then uses the `predict_emb()` function to assign each point in the $p\text{-}D$ dataset to a corresponding hexagon bin in the $2\text{-}D$ model, producing a prediction data frame that contains both the predicted bin assignment (`pred_h`) and the original observation `ID`.

The function joins this prediction output with both the $p\text{-}D$ model and the $p\text{-}D$ data (to retrieve true coordinates). It then calculates squared differences between the original and predicted $p\text{-}D$ coordinates for each dimension, storing these as `error_square_x1`, `error_square_x2`, ..., up to the dimensionality of the data.

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the HBE which is the average of the total squared error per point.

These metrics are returned in a tibble as `Error` (absolute error) and `HBE` (root mean squared error). The computation of total absolute error and HBE is performed in C++ for efficiency using the internal `compute_errors()` function.

@@

Furthermore, `augment()` requires both the fitted model object returned by `fit_highd_model()` and $p\text{-}D$ data to begin. It extends the fitted model by adding prediction results and error diagnostics to the original $p\text{-}D$ data.

The function starts with the same process as is used in the `glance()` function to produce a predicted point in $p\text{-}D$ for each point in the $p\text{-}D$ dataset.

@@

The package offers several $2\text{-}D$ visualizations, including:

@@

The generated $p\text{-}D$ model, overlaid with the data, can also be visualized using `show_langevitour`. Additionally, it features a function for visualizing the $2\text{-}D$ projection of the fitted model overlaid on the data, called `plot_proj`.

@@

The `geom_trimesh()` function introduces a custom `ggplot2` layer designed for visualizing $2\text{-}D$ wireframe on a provided set of bin centroids.

@@

### $p\text{-}D$ model visualization

To visualize how well the $p\text{-}D$ model captures the underlying structure of the high-dimensional data, we provide a tour of the model in $p\text{-}D$ using the `show_langevitour()` function. This function renders a dynamic projection of both the high-dimensional data and the model using the `langevitour` R package [@paul2023].

@@

As an alternative to `langevitour`, users can explore the fitted $p\text{-}D$ model using the `detourr` [@casper2025]. The combined data object from `comb_data_model()` can be passed directly to the `detour()` function, where `tour_aes()` defines the projection variables and color mapping. The visualization is rendered using `show_scatter()`, which can display both data points and the model’s structural edges via the `edges` argument.

@@

In the resulting interactive visualization, blue points represent the high-dimensional data, orange points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped to high-dimensional space.

@@

To support interactive evaluation of how well the $p\text{-}D$ model captures the structure of the high-dimensional data, we introduce `show_link_plots()`. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the $2\text{-}D$ bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), $2\text{-}D$ model data (`model_2d`), and model error (`error_data`).

@@

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

As an alternative to using `langevitour`, link plots can also be generated with the `detourr` package. In this setup, users can manually construct the linked visualization using the `crosstalk` [@joe2023] and `htmltools` [@joe2024] packages. The interactive layout is created by arranging the $2\text{-}D$ NLDR plot (`nldr_plt`), the optional error distribution plot (`error_plt`), and the tour view produced with `detourr`, side by side within a flexible grid. The NLDR and error plots are rendered with `ggplotly()` [@chapman2020] to enable interactive linking. The `bscols()` function from `crosstalk` manages synchronization across these panels, allowing for linked brushing and coordinated selection between these interactive plots.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-hbe) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

error_limb <- error_limb |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  filter(a1 >= 0.03) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup()

@@

error_plot_limb <- plot_hbe(error_limb) +
  scale_x_continuous(breaks = sort(unique(error_limb$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#984ea3','#4daf4a','#ff7f00'))

@@

```{r limb-hbe, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\\label{fig:limb-hbe}", fig.pos='H', echo=TRUE}

@@

plot_hbe_layouts(plots = list(error_plot_limb, nldr1, 
                             nldr2, nldr3, nldr4, 
                             nldr5, nldr6), design = design)
```

@@

```{r model-limb, fig.cap="Compare the published $2\\text{-}D$ layout (Figure \\ref{fig:limb-hbe}b) and the $2\\text{-}D$ layout selected (Figure \\ref{fig:limb-hbe}f) by HBE plot (Figure \\ref{fig:limb-hbe}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", fig.pos='H', fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

@@

The `quollr` package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in $2\text{-}D$ and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a $2\text{-}D$ layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate $2\text{-}D$ representations.

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, `quollr` provides a quantitative route for evaluation. It enables the computation of HBE and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the `langevitour` package, allowing users to inspect where the model fits well and where it does not.

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and $2\text{-}D$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates HBE and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in $2\text{-}D$. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.
